Sequential Composition for Relaxed Memory: Pomsets with Predicate Transformers

### 2021-04-16 paper/intro.tex
intro
repair damage of wordsmithing intro

Introductory programmers are taught \emph{sequential abstraction}: that the
program $\aCmd_1\SEMI \aCmd_2$ executes $\aCmd_1$ before $\aCmd_2$.  Since
the late 1960s, we've been able to explain this using logic
\citep{Hoare:1969:ABC:363235.363259}.  In
\citeauthor{DBLP:journals/cacm/Dijkstra75}'s
[\citeyear{DBLP:journals/cacm/Dijkstra75}] formulation, we think of programs
as \emph{predicate transformers}, where predicates describe the state of
memory in the system.  In the calculus of weakest preconditions, programs map
postconditions to preconditions.  We recall the definition of
$\fwp{\aCmd}{\bForm}$ for loop-free code below (where $\aReg$ ranges over thread-local \emph{registers}
and $\aExp$ ranges over side-effect free \emph{expressions}). 
\begin{multicols}{2}
  \begin{enumerate}[,label=(\textsc{d}\arabic*),ref=\textsc{d}\arabic*]
  \item
    \begin{math}
      \fwp{\SKIP}{\bForm} = \bForm
    \end{math}
    % \item 
    %   \begin{math}
    %     \fwp{\ABORT}{\bForm} = \FALSE
    %   \end{math}
    \stepcounter{enumi}
  %\item[] \labeltext[\textsc{d}2]{}{wp-assign}
  %  \begin{enumerate}[leftmargin=0pt]
  %  \item \label{wp-write}
  %    $\fwp{\PW{\aLoc}{\aExp}}{\bForm} = \bForm[\aExp/\aLoc]$
    \item \label{wp-let}
      $\fwp{\LET{\aReg}{\aExp}}{\bForm} = \bForm[\aExp/\aReg]$
  %  \item \label{wp-read}
  %    $\fwp{\PR{\aLoc}{\aReg}}{\bForm} = \aLoc{=}\aReg\limplies\bForm$ %\bForm[\aLoc/\aReg]$ %
  %  \end{enumerate}
    % \item 
    %   \begin{math}
    %     \fwp{\LET{x}{\aExp}}{\bForm} = \bForm[\aExp/x]
    %   \end{math}
  \item
    \begin{math}
      \fwp{\aCmd_1;\aCmd_2}{\bForm} = \fwp{\aCmd_1}{\fwp{\aCmd_2}{\bForm}}
    \end{math}
  \item
    \begin{math}
      \fwp{\IF{\aExp}\THEN \aCmd_1\ELSE \aCmd_2\FI}{\bForm}= {}
    \end{math}
    \\
    \begin{math}
      ((\aExp{\ne}0) \limplies \fwp{\aCmd_1}{\bForm}) \land
      ((\aExp{=}0) \limplies \fwp{\aCmd_2}{\bForm})
    \end{math}
  \end{enumerate}
\end{multicols}
For this language, the Hoare triple
$\hoare{\aForm}{\aCmd}{\bForm}$ holds exactly when $\aForm \limplies
\fwp{\aCmd}{\bForm}$.
%% We have split
%% \citeauthor{DBLP:journals/cacm/Dijkstra75}'s rule for assignment
%% \eqref{wp-assign} into three cases. In our notation,
%% $\aReg$--$\bReg$ range over thread-local registers, which may be assigned at
%% most once,
%% $\aLoc$--$\cLoc$ range over shared memory references, and
%% $\aExp$--$\bExp$ range over thread-local expressions, which do \emph{not}
%% include
%% $\aLoc$--$\cLoc$.\footnote{Under these assumptions, \eqref{wp-read} is
%%   equivalent to $\fwp{\PR{\aLoc}{\aReg}}{\bForm} = \bForm[\aLoc/\aReg]$.}
%%
This is an elegant explanation of sequential computation in a sequential
context.  In a concurrent context, however, generalizing
\ref{wp-let} from thread-local registers to cross-thread memory locations
is unsound! \ref{wp-let} is sound because a read from a thread-local
register must be fulfilled by a preceding write
in the same thread.  In a concurrent context, aread from a cross-thread
memory location may be fulfilled by a write
in another thread.

@@

Predicate transformers are a good fit for logical models of dependency
calculation, since both are concerned with preconditions and how they are
transformed by sequential composition. Our first attempt is to associate a
predicate transformer with each pomset. We visualize this in diagrams by
showing how $\bForm$ is transformed, for example:
\begin{align*}
  \begin{gathered}
    \PR{x}{r}
    \\
    \hbox{\begin{tikzinline}[node distance=1ex and 1.5em]
        \event{rx0}{\DR{x}{0}}{}
        \xform{rx0d}{(0{=}r) \limplies \bForm}{right=of rx0}
        \xo{rx0}{rx0d}
      \end{tikzinline}}
  \end{gathered}
  &&
  \begin{gathered}
    \PR{y}{s}
    \\
    \hbox{\begin{tikzinline}[node distance=1ex and 1.5em]
        \event{ry0}{\DR{y}{0}}{}
        \xform{ry0d}{(0{=}s) \limplies \bForm}{right=of ry0}
        \xo{ry0}{ry0d}
      \end{tikzinline}}
  \end{gathered}
  &&
  \begin{gathered}
    \IF{s{<}1} \THEN \PW{z}{r{*}s} \FI
    \\
    \hbox{\begin{tikzinline}[node distance=1ex and 1.5em]
        \event{wz0}{(s {<} 1) \land (r{*}s){=}0 \mid \DW{z}{0}}{}
        \xform{wz0d}{\bForm[r{*}s/z]}{right=of wz0}
        \xo{wz0}{wz0d}
      \end{tikzinline}}
  \end{gathered}
\end{align*}
The predicate transformer from the write matches
\citeauthor{DBLP:journals/cacm/Dijkstra75}.  For the reads, however,
\ref{wp-read} defines the transformer of $\PR{x}{r}$ to be
$(x{=}r) \limplies \bForm$.  Instead, we use $(0{=}r) \limplies \bForm$,
reflecting the fact that $0$ may come from a concurrent write.  The
obligation to find a matching write is moved from the sequential semantics of
\emph{substitution} and \emph{implication} to the concurrent semantics of
\emph{fulfillment}.

@@

We show how predicate transformers~\cite{DBLP:journals/cacm/Dijkstra75} can
be added to pomsets with
preconditions~\cite{DBLP:journals/pacmpl/JagadeesanJR20} to create a
compositional semantics for sequential composition.
\begin{itemize}
\item \S\ref{sec:model} presents the basic model, which satisfies many
  desiderata, but not all.
\item \S\ref{sec:arm} shows two approaches for efficient implementation on
  Arm.  The first uses a suboptimal lowering for acquiring reads.  The second
  uses an optimal lowering, but requires changes to the definitions of
  parallel and sequential composition.
\item \S\ref{sec:variants} generalizes the basic semantics of read and write
  to validate compiler optimizations.
\end{itemize}
% Acknowledgements go here, once we're not double-blinded.
% The definitions in this paper have been formalized in Agda.
Because it is closely related, we expect that the memory-model results of
\cite{DBLP:journals/pacmpl/JagadeesanJR20} apply to our model, including
compositional reasoning for temporal safety properties and {local} \drfsc{}
as in \cite{Dolan:2018:BDR:3192366.3192421,DBLP:conf/ppopp/DongolJR19,promising-ldrf}.
% In \textsection\ref{sec:arm}, we provide an alternative proof strategy for
% efficient compilation to \armeight{}, which improves upon that of
% \cite{DBLP:journals/pacmpl/JagadeesanJR20} by using a recent alternative
% characterization of \armeight{}.

>>>>

Introductory programmers are taught \emph{sequential abstraction}: that the
program $\aCmd_1\SEMI \aCmd_2$ executes $\aCmd_1$ before $\aCmd_2$.  Since
the late 1960s, we've been able to explain this using logic
\citep{Hoare:1969:ABC:363235.363259}.  In
\citeauthor{DBLP:journals/cacm/Dijkstra75}'s
[\citeyear{DBLP:journals/cacm/Dijkstra75}] formulation, we think of programs
as \emph{predicate transformers}, where predicates describe the state of
memory in the system.  In the calculus of weakest preconditions, programs map
postconditions to preconditions.  We recall the definition of
$\fwp{\aCmd}{\bForm}$ for loop-free code below (where $\aReg$--$\bReg$ range over thread-local \emph{registers}
and $\aExp$--$\bExp$ range over side-effect free \emph{expressions}). 
%\begin{multicols}{2}
  \begin{enumerate}[,label=(\textsc{d}\arabic*),ref=\textsc{d}\arabic*]
  \item
    \begin{math}
      \fwp{\SKIP}{\bForm} = \bForm
    \end{math}
  \item \label{wp-let}
    $\fwp{\LET{\aReg}{\aExp}}{\bForm} = \bForm[\aExp/\aReg]$
    % \item 
    %   \begin{math}
    %     \fwp{\ABORT}{\bForm} = \FALSE
    %   \end{math}
    %   \stepcounter{enumi}
    % \item[] \labeltext[\textsc{d}2]{}{wp-assign}
    %   \begin{enumerate}[leftmargin=0pt]
    %   \item \label{wp-write}
    %     $\fwp{\PW{\aLoc}{\aExp}}{\bForm} = \bForm[\aExp/\aLoc]$
    %   \item \label{wp-let}
    %     $\fwp{\LET{\aReg}{\aExp}}{\bForm} = \bForm[\aExp/\aReg]$
    %   \item \label{wp-read}
    %     $\fwp{\PR{\aLoc}{\aReg}}{\bForm} = \aLoc{=}\aReg\limplies\bForm$ %\bForm[\aLoc/\aReg]$ %
    %   \end{enumerate}
    % \item 
    %   \begin{math}
    %     \fwp{\LET{x}{\aExp}}{\bForm} = \bForm[\aExp/x]
    %   \end{math}
  \item
    \begin{math}
      \fwp{\aCmd_1;\aCmd_2}{\bForm} = \fwp{\aCmd_1}{\fwp{\aCmd_2}{\bForm}}
    \end{math}
  \item
    \begin{math}
      \fwp{\IF{\aExp}\THEN \aCmd_1\ELSE \aCmd_2\FI}{\bForm}= {}
    \end{math}
    \begin{math}
      ((\aExp{\ne}0) \limplies \fwp{\aCmd_1}{\bForm}) \land
      ((\aExp{=}0) \limplies \fwp{\aCmd_2}{\bForm})
    \end{math}
  \end{enumerate}
%\end{multicols}
For this language, the Hoare triple
$\hoare{\aForm}{\aCmd}{\bForm}$ holds exactly when $\aForm \limplies
\fwp{\aCmd}{\bForm}$.
%% We have split
%% \citeauthor{DBLP:journals/cacm/Dijkstra75}'s rule for assignment
%% \eqref{wp-assign} into three cases. In our notation,
%% $\aReg$--$\bReg$ range over thread-local registers, which may be assigned at
%% most once,
%% $\aLoc$--$\cLoc$ range over shared memory references, and
%% $\aExp$--$\bExp$ range over thread-local expressions, which do \emph{not}
%% include
%% $\aLoc$--$\cLoc$.\footnote{Under these assumptions, \eqref{wp-read} is
%%   equivalent to $\fwp{\PR{\aLoc}{\aReg}}{\bForm} = \bForm[\aLoc/\aReg]$.}
%%
This is an elegant explanation of sequential computation in a sequential
context. Note that \ref{wp-let} is sound because a read from a thread-local
register must be fulfilled by a preceding write in the same thread.
In a concurrent context, with shared variables
($\aLoc$--$\cLoc$), the obvious generalizations
\begin{multicols}{2}
\begin{itemize}
\item[{\labeltext[\textsc{d}2b]{(\textsc{d}2b)}{wp-write}}] $\fwp{\LET{\aLoc}{\aExp}}{\bForm} = \bForm[\aExp/\aLoc]$
\item[{\labeltext[\textsc{d}2c]{(\textsc{d}2c)}{wp-read}}] $\fwp{\LET{\aReg}{\aLoc}}{\bForm} = \bForm[\aLoc/\aReg]$
\end{itemize}
\end{multicols}
\noindent
are unsound!  In particular, a read from a shared memory location may be fulfilled by a write
in another thread, invalidating \ref{wp-read}.  (We assume that expressions
% $\aExp$--$\bExp$
do \emph{not} include shared variables.) % $\aLoc$--$\cLoc$.)

@@

Predicate transformers are a good fit for logical models of dependency
calculation, since both are concerned with preconditions and how they are
transformed by sequential composition. Our first attempt is to associate a
predicate transformer with each pomset. We visualize this in diagrams by
showing how $\bForm$ is transformed, for example:
\begin{align*}
  \begin{gathered}
    \PR{x}{r}
    \\
    \hbox{\begin{tikzinline}[node distance=1ex and 1.5em]
        \event{rx0}{\DR{x}{0}}{}
        \xform{rx0d}{(0{=}r) \limplies \bForm}{right=of rx0}
        \xo{rx0}{rx0d}
      \end{tikzinline}}
  \end{gathered}
  &&
  \begin{gathered}
    \PR{y}{s}
    \\
    \hbox{\begin{tikzinline}[node distance=1ex and 1.5em]
        \event{ry0}{\DR{y}{0}}{}
        \xform{ry0d}{(0{=}s) \limplies \bForm}{right=of ry0}
        \xo{ry0}{ry0d}
      \end{tikzinline}}
  \end{gathered}
  &&
  \begin{gathered}
    \IF{s{<}1} \THEN \PW{z}{r{*}s} \FI
    \\
    \hbox{\begin{tikzinline}[node distance=1ex and 1.5em]
        \event{wz0}{(s {<} 1) \land (r{*}s){=}0 \mid \DW{z}{0}}{}
        \xform{wz0d}{\bForm[r{*}s/z]}{right=of wz0}
        \xo{wz0}{wz0d}
      \end{tikzinline}}
  \end{gathered}
\end{align*}
The predicate transformer from the write matches
\citeauthor{DBLP:journals/cacm/Dijkstra75}'s \ref{wp-write}.  For the reads,
however, \ref{wp-read} defines the transformer of $\PR{x}{r}$ to be
$\bForm[x/r]$, which is equivalent to $(x{=}r) \limplies \bForm$ under the
assumption that registers are assigned at most once.
Instead, we use $(0{=}r) \limplies \bForm$, reflecting the fact that $0$ may
come from a concurrent write.  The obligation to find a matching write is
moved from the sequential semantics of \emph{substitution} and
\emph{implication} to the concurrent semantics of \emph{fulfillment}.

@@

We show how predicate transformers~\cite{DBLP:journals/cacm/Dijkstra75} can
be added to pomsets with
preconditions~\cite{DBLP:journals/pacmpl/JagadeesanJR20} to create a
compositional semantics for sequential composition.
\begin{itemize}
\item \S\ref{sec:model} presents the basic model, which satisfies many
  desiderata, but not all.
\item \S\ref{sec:arm} shows two approaches for efficient implementation on
  Arm.  The first uses a suboptimal lowering for acquiring reads.  The second
  uses an optimal lowering, but requires changes to the definitions of
  sequential composition.
\item \S\ref{sec:variants} generalizes the basic semantics of read and write
  to validate compiler optimizations.
\end{itemize}
% Acknowledgements go here, once we're not double-blinded.
% The definitions in this paper have been formalized in Agda.
Because it is closely related, we expect that the memory-model results of
\cite{DBLP:journals/pacmpl/JagadeesanJR20} apply to our model, including
compositional reasoning for temporal safety properties and {local} \drfsc{}
as in \cite{Dolan:2018:BDR:3192366.3192421,DBLP:conf/ppopp/DongolJR19,promising-ldrf}.
% In \textsection\ref{sec:arm}, we provide an alternative proof strategy for
% efficient compilation to \armeight{}, which improves upon that of
% \cite{DBLP:journals/pacmpl/JagadeesanJR20} by using a recent alternative
% characterization of \armeight{}.
